Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-04-09 ms.tex
various

An important caveat, however, is that $\vw$ is only a few times higher than the stellar velocity dispersion, i.e. $\zeta \sim \zeta_c$ (eq.~[\ref{eq:zetacrit}]).  This implies that the true stagnation radius and SMBH accretion rate could be larger than our analytic estimates, potentially resulting in thermal instability, especially for core galaxies with large break radii ($\S\ref{sec:mdot}$).

@@

  Perhaps a more readily addressable test is whether a steady-state
  accretion picture developed in this work can account for the the
  large scatter, typically of $2-3$ orders of magnitude, in
  $L_{X}/L_{\rm edd}$ at fixed $M_{\bullet}$.  Scatter could result
  from at least three effects: (1) intrinsic scatter in the scaling
  relationships ($\Mbh-\sigma$ and $\Mbh-L_{\rm bulge}$) used to
  measure $\Mbh$; (2) the strong sensitivity of the accretion rate to
  the stellar wind velocity and mass loss parameter,
  $\dot{M}/\dot{M}_{\rm edd} \propto \eta v_{w}^{-3.8(-2.4)}$ for
  core(cusp) galaxies, respectively (Fig.~\ref{fig:mdot_mass};
  eq.~[\ref{eq:eddr_analytic}]).  When combined with the significant
  dependence of $v_w$ on stochastic intermittency in the star
  formation history (Appendix \ref{fig:NickPlot2}), this can lead to
  order of magnitude differences in $\dot{M}$; (3) differences in angular
  momentum of the infalling gas from galaxy to galaxy, resulting in
  differences in the fraction of the gas lost to outflows.
  Also potentially contributing is the order of magnitude difference,
  at fixed $v_w$ and $M_{\bullet}$, between $\dot{M}$ for core and
  cusp galaxies (Fig.~\ref{fig:bh_growth}).  Differences in $\dot{M}$
  are augmented by the expectation that $L_{\rm X} \propto \dot{M}^{2}
  $ for radiatively inefficient flows.

@@

It is also possible that many of the \citet{Miller+15} sample of
galactic nuclei are not accreting in steady state.  The 
X-ray efficiency assumed in the right panel of Figure \ref{fig:miller}
represents a maximum value, which does not for instance account for
suppression of the SMBH accretion rate due to outflows from the
accretion disk on small radial scales (e.g.,
\citealt{Blandford&Begelman99}).  Furthermore, our SFH-averaged
accretion rates are derived using analytic expressions for the
stagnation radius that may underpredict the true accretion rate if
$\zeta < \zeta_c$, as may occur in massive core galaxies with large
break radii.  If the actual radiative efficiency is lower than our
fiducial assumption, then many of the X-ray detections would lie above
the predictions for stable accretion, i.e., $L_X \gtrsim \epsilon_X \dot{M}_{\rm TI}$ (yellow line).  In this case the X-ray detected galaxies experience
less heating than average galaxies of the same mass, and may not be
stably accreting at all.  This is particularly plausible because the
\citet{Miller+15} sample includes only early-type galaxies, with older
stellar populations than the average for their black hole mass.

@@

\item Radiative cooling has a more pronounced influence on the flow structure.  Cooling instabilities are likely to develop when the radiative cooling rate exceeds the gas heating rate near the stagnation radius.  This condition is approximately equivalent to the more criterion $t_{\rm cool} \gg t_{\rm ff}$ criterion advocated by \citet{McCourt+12} and \citet{Li&Bryan14a}.

@@

\item Only for high X-ray radiative efficiencies similar to the maximum values predicted by ADAF models are the inferred accretion rates of quiescent early-type galaxies below the maximum value allowed for thermally-stable accretion $\dot{M}_{\rm TI}$ (Fig.~\ref{fig:miller}).  It is thus well possible that the current sample of X-ray detections is instead comprised mainly of nuclei undergoing non-steady accretion due to (local) thermal instabilities.  In this case there may exist a separate branch of low $L_x$ (undetected) thermally-stable nuclei, especially at low $M_{\bullet}$.  Such a bimodal population would call into question extrapolating the $L_{\rm X}(M_{\bullet}$) relationship to low $M_{\bullet}$ to constrain the SMBH occupation fraction in low mass galaxies (\citealt{Miller+15}).

@@

We conclude by drawing attention to the following limitations of our model
    \begin{enumerate}
    \item We assume that there is no large scale inflow of gas to the
      nucleus, and the only source of gas feeding the black hole is
      stellar wind mass loss from the local stellar population. Thus,
      our estimates of gas density are a lower bound on the nuclear
      gas density {\bf AG: make sense?}.

      As we have already pointed out AGN activity cannot be explained
      by steady thermally stable accretion of mass lost from stellar
      winds.

    \item Overall the accretion rate onto the central SMBH will be
      strongly tied to the properties of the stellar population
      (e.g. age) within the sphere of influence of the black
      hole. Thus, our model cannot say anything definitive about the
      mode accretion for individual galaxies unless something is known
      about their specific star formation history.

    \item We adopt halo average star formation rates measured in
      cosmological simulations by \citet{MosterNaab+:2013a}, and
      neglect any spatial variation in the star formation rate for any
      given galaxy. Galaxies with masses$\gsim 7\times 10^{10} \Msun$,
      assemble their stars from the inside out
      \citep{PerezCidFernandes+:2013a} and so for a given halo mass
      the galactic nucleus may be significantly older than in our
      model and thus more prone to thermal instability.

      On the other hand, \citealt{LauerFaber+:2005a} find that $\sim
      60\%$ of cusp galaxies and $\sim29\%$ of core galaxies have
      (generally unresolved) emission in excess of the inward
      extrapolation of the Nuker law, and noted that these nuclei are bluer
      than the background galaxy starlight, suggesting the presence of
      younger stars towards the center of the galaxy {\bf AG mass
        threshold for this effect}.

      In fact it is possible that the mass injection is dominated by a
      nuclear star cluster with a break radius on parsec
      scales. Although this is much smaller than our fiducial break
      radius (100 pc). If the heating is sufficiently high moving the
      break radius inwards would merely change the outer gas density
      profile, without changing the stagnation radius in the flow and
      thus the accretion rate onto the black hole. Of course if one
      moves the break radius inside where the stagnation radius would
      be for large break radii, the stagnation radius would move
      inwards somewhat. However note the for high heating rate ($v_w
      \gsim 500 $ km/s and small black hole masses $\Mbh \lsim 10^7
      \Msun$), we predict stagnation radii smaller than a parsec.

>>>>

An important caveat, however, is that $\vw$ is only a few times higher than the stellar velocity dispersion, i.e. $\zeta \sim \zeta_c$ (eq.~[\ref{eq:zetacrit}]).  This implies that the true stagnation radius and SMBH accretion rate could be larger than our analytic estimates, potentially resulting in thermal instability, especially for core galaxies with large break radii ($\S\ref{sec:mdot}$).  In other words, the CNM of some subset of massive ellipticals be thermally unstable, not necessarily because they are receiving lower stellar feedback than the average for their galaxy mass, but due to the high heating required for stability given the structure of their stellar potential.

@@

  Perhaps a more readily addressable test is whether a steady-state
  accretion picture developed in this work can account for the the
  large scatter, typically of $2-3$ orders of magnitude, in
  $L_{X}/L_{\rm edd}$ at fixed $M_{\bullet}$.  Scatter could result
  from at least three effects: (1) intrinsic scatter in the scaling
  relationships ($\Mbh-\sigma$ and $\Mbh-L_{\rm bulge}$) used to
  measure $\Mbh$; (2) the strong sensitivity of the accretion rate to
  the stellar wind velocity and mass loss parameter,
  $\dot{M}/\dot{M}_{\rm edd} \propto \eta v_{w}^{-3.8(-2.4)}$ for
  core(cusp) galaxies, respectively (Fig.~\ref{fig:mdot_mass};
  eq.~[\ref{eq:eddr_analytic}]).  When combined with the significant
  dependence of $v_w$ on stochastic intermittency in the star
  formation history (Appendix \ref{fig:NickPlot2}), this can lead to
  order of magnitude differences in $\dot{M}$; (3) differences in angular
  momentum of the infalling gas from galaxy to galaxy, resulting in
  differences in the fraction of the gas lost to outflows.
  Also potentially contributing is the order of magnitude difference,
  at fixed $v_w$ and $M_{\bullet}$, between $\dot{M}$ for core and
  cusp galaxies (Fig.~\ref{fig:bh_growth}).  Differences in $\dot{M}$
  are augmented by the theoretical expectation that $L_{\rm X} \propto \dot{M}^{2}
  $ for radiatively inefficient flows.

@@

It is also possible that many of the \citet{Miller+15} sample of
galactic nuclei are not accreting in steady state.  The 
X-ray efficiency assumed in the right panel of Figure \ref{fig:miller}
represents a maximum value, which does not for instance account for
suppression of the SMBH accretion rate due to outflows from the
accretion disk on small radial scales (e.g.,
\citealt{Blandford&Begelman99}).  Furthermore, our SFH-averaged
accretion rates are derived using analytic expressions for the
stagnation radius that may underpredict the true accretion rate if
$\zeta < \zeta_c$, as may occur in massive core galaxies with large
break radii.

If the actual radiative efficiency is lower than our
fiducial assumption, then many of the X-ray detections would lie above
the predictions for stable accretion, i.e., $L_X \gtrsim \epsilon_X \dot{M}_{\rm TI}$ (yellow line).  In this case the X-ray detected galaxies experience
less heating than average galaxies of the same mass, and may not be
stably accreting at all.  This is particularly plausible because the
\citet{Miller+15} sample includes only early-type galaxies, with older
stellar populations than the average for their black hole mass.

@@

\item Radiative cooling has a more pronounced influence on the flow structure.  Cooling instabilities are likely to develop when the radiative cooling rate exceeds the gas heating rate near the stagnation radius.  This condition is approximately equivalent to the criterion $t_{\rm cool} \gg t_{\rm ff}$ criterion advocated by \citet{McCourt+12} and \citet{Li&Bryan14a}.

@@

\item Only for high X-ray radiative efficiencies similar to the maximum values predicted by ADAF models are the inferred accretion rates of quiescent early-type galaxies below the maximum value allowed for thermally-stable accretion $\dot{M}_{\rm TI}$ (Fig.~\ref{fig:miller}).  It is thus well possible that the current sample of X-ray detections is instead comprised mainly of nuclei undergoing non-steady accretion due to (local) thermal instabilities.  In this case there may exist a separate branch of low $L_X$ (undetected) thermally-stable nuclei, especially at low $M_{\bullet}$.  The existence of such a bimodal population would call into question extrapolating the $L_{\rm X}(M_{\bullet}$) relationship to low $M_{\bullet}$ to constrain the SMBH occupation fraction in low mass galaxies (\citealt{Miller+15}).

@@

We conclude by drawing attention to the following limitations of our model.  First, by neglecting any large scale inflow of gas to the nucleus and assuming the only source of gas feeding the black hole is stellar wind mass loss from the local stellar population, our estimates represent a lower bound on the gas density of the CNM.  Our model therefore cannot account for most AGN activity.

The predicted SMBH accretion rate is also strongly tied to the age and radial distribution of the stellar population within the sphere of influence.  Our model cannot make definitive predictions for the accretion mode of individual galaxies without knowledge of their specific SFH.

We adopt halo average star formation rates measured in cosmological simulations by \citet{MosterNaab+:2013a}, and neglect any spatial variation in the star formation rate for any given galaxy. Galaxies of mass $\gsim 7\times 10^{10} \Msun$, assemble their stars from the inside out \citep{PerezCidFernandes+:2013a}, such that for a given halo mass the galactic nucleus may be older than in our model and thus more prone to thermal instability.  On the other hand, \citealt{LauerFaber+:2005a} find that $\sim
      60\%$ of cusp galaxies and $\sim29\%$ of core galaxies have
      (generally unresolved) emission in excess of the inward
      extrapolation of the Nuker law.  They note that these nuclei are bluer
      than the background galaxy starlight, suggesting the presence of
      younger stars towards galactic centre {\bf AG mass
        threshold for this effect}.

      In fact it is possible that the mass injection is dominated by a
      nuclear star cluster with a break radius on parsec
      scales, i.e. much smaller than our fiducial break
      radius of $r_b \sim 100$ pc. If the heating is sufficiently high moving the
      break radius inwards would merely change the outer gas density
      profile, without changing the stagnation radius in the flow and
      thus the accretion rate onto the black hole. Of course if one
      moves the break radius inside where the stagnation radius would
      be for large break radii, the stagnation radius would move
      inwards somewhat. However note the for high heating rate ($v_w
      \gsim 500 $ km/s and small black hole masses $\Mbh \lsim 10^7
      \Msun$), we predict stagnation radii smaller than a parsec.
